Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-11-14 tex/manuscript.tex
Merged revisions

%\documentclass{bio}
\documentclass[oupdraft]{bio}
\usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

@@

%% NOTES TO SELF:
% Cite using:
%    \citet{...}  in text
%    \citealp{...}   within parantheses
%
% Temporary notifications and reminders:
%    \COMMENT{text}{comment to text}
%    \NB{comment (as separate paragraph)}
%    \NOTE{comment inside paragraph}
%    \CITE{missing/unspecified citation}
%\usepackage{color}
%\definecolor{NBcol}{rgb}{0.7,0,1}
%\definecolor{CITEcol}{rgb}{1,0,0.7}
%\definecolor{COMMENTcol}{rgb}{0.7,0.2,0}
%\newcommand\COMMENT[2]{\textcolor{COMMENTcol}{#1}\NOTE{#2}}
%\newcommand\NB[1]{\textcolor{NBcol}{\textit{#1}}}
%\newcommand\NOTE[1]{\NB{[#1]}}
%\newcommand\CITE[1]{\textcolor{CITEcol}{[#1]}}

@@

% Manuscript version (date) in separate footnote
%\renewcommand{\thefootnote}{§}\footnotetext{\color{blue}Manuscript version \today.}

@@

\begin{abstract}
{Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data. In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects. Many methods and tools exist for this purpose. One method, ComBat, which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous. It also has the option of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments. Unfortunately, this frequently used and recommended approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner. The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.
}
\end{abstract}

@@

A popular tool for batch adjustment of gene expression data is ComBat (\citealp{Johnson2007}) which is now included in the R bioconductor package \texttt{sva} (\citealp{Leek2012}). This allows the inclusion of covariates, e.g. group difference, the effects of which should not be removed in the batch adjustment. In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches. Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora. ComBat has thus helped popularise this approach. Based on the actual use of ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported. The commercial software Partek (Partek® Genomics Suite®, version 6.6 Copyright ©; 2014 Partek Inc., St. Louis, MO, USA), and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{http://link.springer.com/article/10.1007/s12561-013-9081-1}) also offer batch adjustment with covariates. Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that the this is not intended used prior to linear modeling, we choose to focus more on the popular and well documented ComBat.

@@

@@

@@

@@

If we estimate the group difference within batch $i$, we get
\begin{equation}
\Delta\hat{\beta}_i=\bar{Y}_{iA}-\bar{Y}_{iB}\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu_i}\right)
\quad\text{where}\quad
\bar{Y}_{ij}=\frac{\sum_{r=1}^{n_{ij}}Y_{ijr}}{n_{ij}},\,
\nu_i=\frac{1}{\frac{1}{n_{iA}}+\frac{1}{n_{iB}}},
\end{equation}
from which we may express the overall estimate of $\Delta\beta$
\begin{equation}
\Delta\hat{\beta}=\frac{\sum_{i=1}^m \nu_i\,\Delta\hat{\beta}_i}{\nu}
\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu}\right)
\quad\text{where}\quad
\nu=\sum_{i=1}^m\nu_i.
\end{equation}

If batch and group effects are estimated using a two-way ANOVA, the estimate $\Delta\hat{\beta}$ will be as stated above, and so the estimated group difference is unaffected. The batch effects are then removed, and the estimated group differences retained, leaving the estimated $\Delta\hat{\beta}$ unchanged by the batch adjustment. However, if this batch adjusted data set is analysed without considering batch effects, the variance of $\Delta\hat{\beta}$ will be computed under the assumption that it is derived from a comparison of $n_A=\sum_{i=1}^m n_{iA}$ versus $n_B=\sum_{i=1}^m n_{iB}$ samples, and thus satisfy
\begin{equation}
\Delta\hat{\beta}
\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu_0}\right)
\quad\text{where}\quad
\nu_0=\frac{1}{\frac{1}{n_A}+\frac{1}{n_B}}.
\end{equation}
Using Jensen's inequality, we may derive
\begin{equation}
\nu_0
=\frac{1}{\frac{1}{n_A}+\frac{1}{n_B}}=n\cdot\frac{n_A}{n}\cdot\frac{n_B}{n}
\ge
\nu
=\sum_{i=1}^m \frac{1}{\frac{1}{n_{iA}}+\frac{1}{n_{iB}}}
=\sum_{i=1}^m n_i\cdot\frac{n_{iA}}{n_i}\cdot\frac{n_{iB}}{n_i}
\end{equation}
with equality if and only if the ratios $n_{iA}:n_{iB}=n_A:n_B$ for all batches $i=1,\ldots,m$.

@@

The extent to which batch adjustment with group differences retained will confound subsequent analyses depends on the batch--group balance. We have reanalysed two cases with varying degree of unbalance.

@@

In an experiment described in \citet{Towfic2014}, the effect of glatiramer acetate (a medicine for multiple sclerosis) was compared to the effect of a generic drug. Cells were treated with
glatiramer acetate (34 samples), the generic drug (11 samples) or 14 different types of other treatments(64 samples), and mRNA was measured using microarrays. A batch effect correlating to the chip (Illumina
WG-6\_V2, six samples per chip, 17 chips in total) was observed and adjusted for with
ComBat, whereafter the data were tested for differentially expressed
genes, yielding hundreds of differentially expressed probes (Table S5, \citealp{Towfic2014}) using a $\text{FDR}<0.05 threshold.
Unfortunately, the batch--treatment design was unbalanced, with several
batches having only one of the main treatments of interest. When we
re-analyzed the data without the use of ComBat, but instead blocked for
batch effect in \texttt{limma}, only 9 genes were found ($\text{FDR}<0.05$)
. The sanity check  outlined above with random numbers
was also carried out. The distribution of p-values for different settings
are shown in Figure~\ref{fig:comparison}a. Our conclusion is that most of
the genes found as differentially expressed in the analysis described in
\citet{Towfic2014} are false positives. This example represents somewhat
of a ``worst case'' scenario for applying ComBat, since it both has a
very unbalanced batch--group design and an a priori assumption of no
difference. Our re-analysis, although based on the "methods" section of
Towfic et al, deviates in important ways from the analysis that was
actually performed in the paper (personal communication, Ben Zeskind).
More detailed information on this aspect can be found in the extended github repostiory

@@

A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were applied, i.e batch labels or whether group labels were supplied as covariates. Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches. The incorporation of ComBat into various analysis tools appears to make it more accessible, but its use less transparent. Such tools include GenePattern (v3.9.0) (\citealp{Reich2006}), AltAnalyze (v2.0.8 http://nar.oxfordjournals.org/content/38/suppl_2/W755.full) and SCAN.UPC (v2.6.3 http://www.pnas.org/content/110/44/17778.long), which offers ComBat with covariates, as well as TCGA and  inSilicoMerging (v1.8.7) (\citealp{ http://www.biomedcentral.com/1471-2105/13/335}), presently without covariates as an option. Especially troublesome, are the ComBat implementations in ChAMP (v1.2.8) (\citealp{http://bioinformatics.oxfordjournals.org/content/30/3/428.full}) and intCor (v1.03) (http://bioinformatics.oxfordjournals.org/content/early/2014/07/24/bioinformatics.btu423.full), where the use of covariates seems automatic and hidden from the user in the method call, while there is no mention of covariates in the documentation or articles.

@@

\bibliographystyle{biorefs}
\input{manuscript.bbl} % Include bibliography bbl-file
%\bibliography{D:/Projects/_Appl_/_TeX_/library} % To use with BibTeX to generate bbl-file
%\bibliography{refs}

@@

\begin{figure}[!p]
\begin{center}$
\begin{array}{cc}
\includegraphics[width=4cm]{Fig/leekrandomdatapvalues.pdf} &
\includegraphics[width=4cm]{Fig/leekqqplot.png}\includegraphics[width=4cm]{Fig/samplesizescaling_ComBat.pdf}
\end{array}$
\end{center}
\caption
{
This is a sanity check where the recommended use of ComBat fails, adapted from the user guide in the \texttt{sva} package. Real data are substituted with random numbers from a normal distribution, but the batch--group design is retained. ComBat is applied, followed by an F-test.
a) P-value distribution
b) QQ plot of the F-statistics
c) P-value distributions for 3 equally unbalanced random number experiments with different sample sizes, 12,120 and 1200 samples from two study groups with a 1:5 and 5:1 distribution in two batches. A random batch effect is added for 10\% of the 20000 genes. This example is not from the \texttt{sva} package.
}
\label{fig:sanity}
\end{figure}

>>>>

\documentclass{bio}
%\documentclass[oupdraft]{bio}
\usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

@@

%% NOTES TO SELF:
% Cite using:
%    \citet{...}  in text
%    \citealp{...}   within parantheses
%
% Temporary notifications and reminders:
%    \COMMENT{text}{comment to text}
%    \NB{comment (as separate paragraph)}
%    \NOTE{comment inside paragraph}
%    \CITE{missing/unspecified citation}
\usepackage{color}
\definecolor{NBcol}{rgb}{0.7,0,1}
\definecolor{CITEcol}{rgb}{1,0,0.7}
\definecolor{COMMENTcol}{rgb}{0.7,0.2,0}
\newcommand\COMMENT[2]{\textcolor{COMMENTcol}{#1}\NOTE{#2}}
\newcommand\NB[1]{\textcolor{NBcol}{\textit{#1}}}
\newcommand\NOTE[1]{\NB{[#1]}}
\newcommand\CITE[1]{\textcolor{CITEcol}{[#1]}}
\newcommand\citeurl[1]{\textcolor{CITEcol}{[\url{#1}]}}

@@

% Manuscript version (date) in separate footnote
\renewcommand{\thefootnote}{§}\footnotetext{\color{blue}Manuscript version: \today.}

@@

\begin{abstract}
{Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data. In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects. Many methods and tools exist for this purpose. One method, ComBat, which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous. It also has the option of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments. This is a natural approach when study groups are not balanced between batches as batch adjustments may otherwise bias, usually deflate, group differences. Unfortunately, this frequently used and recommended approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner. \COMMENT{The scientific community seems to be largely unaware of how this approach may lead to false discoveries.}{Har moderert påstanden noe, jfr. reviewer~2, pt.~6.} We would expect any method for removing batch effects while retaining group differences to run into the same problem.
}
\end{abstract}

@@

A popular tool for batch adjustment of gene expression data is ComBat (\citealp{Johnson2007}) which is now included in the R bioconductor package \texttt{sva} (\citealp{Leek2012}). This allows the inclusion of covariates, e.g. group difference, the effects of which should not be removed in the batch adjustment. In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches. Increasingly, the inclusion of group difference as a covariate when removing batch effects has previously been recommended, both in the \texttt{sva} tutorial and user fora. ComBat has thus helped popularise this approach. Based on the actual use of ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported. The commercial software Partek \CITE{citation or reference needed?}, and the R packages \texttt{limma} (\citealp{Smyth2003}) and \texttt{ber} (\citealp{Giordan2013}) also offer batch adjustment with covariates. Seeing that \texttt{ber} is sparsly used, and \texttt{Partek} and the relevant method ``removeBatchEffect'' in \texttt{limma} note that the this is not intended used prior to linear modeling, we choose to use the more popular and well documented ComBat to demonstrate the problem.

@@

In conclusion, when study groups are unevenly distributed across batches, batch effects can have confounding effects if not properly adjusted for. This is difficult to do through a two-step procedure as batch adjustment will not be able to produce a ``batch effect free'' data set, as has also been observed in \citet{Buhule2014}. Estimating batch and group effects simultaneously through two-way ANOVA tries to overcome this, and does succeed at removing systematic biases in the group differences induced by the batch effects, but at the cost of making the observations interdependent to an extent that may invalidate subsequent analyses. Proper statistical analysis of unbalanced designs require models that handle batch effects as part of the analysis. However, the best way to handle this will be to ensure a balanced study design from the start, both to avoid data analysis problems and the loss of statistical power the ensues when batch and group effects need to be disentangled.

@@

\subsubsection{Zero-centering batches:}

@@

\subsubsection{Batch adjustment using two-way ANOVA to estimate batch effects:}

@@

\subsubsection{Group comparison from two-way ANOVA:}

If we estimate the group difference within batch $i$, we get
\begin{equation}
\Delta\hat{\beta}_i=\bar{Y}_{iA}-\bar{Y}_{iB}\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu_i}\right)
\quad\text{where}\quad
\bar{Y}_{ij}=\frac{\sum_{r=1}^{n_{ij}}Y_{ijr}}{n_{ij}},\,
\nu_i=\frac{1}{\frac{1}{n_{iA}}+\frac{1}{n_{iB}}},
\end{equation}
from which we may express the overall estimate of $\Delta\beta$
\begin{equation}
\Delta\hat{\beta}=\frac{\sum_{i=1}^m \nu_i\,\Delta\hat{\beta}_i}{\nu}
\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu}\right)
\quad\text{where}\quad
\nu=\sum_{i=1}^m\nu_i.
\end{equation}
This is the same as would be obtained from a two-way ANOVA analysis.

\subsubsection{Group comparison from one-way ANOVA after batch adjustment:}

If batch and group effects are estimated using a two-way ANOVA, the estimate $\Delta\hat{\beta}$ will be as stated above, and so the estimated group difference is unaffected. The batch effects are then removed, and the estimated group differences retained, leaving the estimated $\Delta\hat{\beta}$ unchanged by the batch adjustment.

However, if this batch adjusted data set is analysed without considering batch effects, the variance of $\Delta\hat{\beta}$ will be computed under the assumption that it is derived from a comparison of $n_A=\sum_{i=1}^m n_{iA}$ versus $n_B=\sum_{i=1}^m n_{iB}$ samples, and thus satisfy
\begin{equation}
\Delta\hat{\beta}
\sim\text{N}\left(\Delta\beta,\frac{\sigma^2}{\nu_0}\right)
\quad\text{where}\quad
\nu_0=\frac{1}{\frac{1}{n_A}+\frac{1}{n_B}}.
\end{equation}
Using Jensen's inequality, we may derive
\begin{equation}
\nu_0
=\frac{1}{\frac{1}{n_A}+\frac{1}{n_B}}=n\cdot\frac{n_A}{n}\cdot\frac{n_B}{n}
\ge
\nu
=\sum_{i=1}^m \frac{1}{\frac{1}{n_{iA}}+\frac{1}{n_{iB}}}
=\sum_{i=1}^m n_i\cdot\frac{n_{iA}}{n_i}\cdot\frac{n_{iB}}{n_i}
\end{equation}
with equality if and only if the ratios $n_{iA}:n_{iB}=n_A:n_B$ for all batches $i=1,\ldots,m$.

@@

The extent to which batch adjustment with group differences retained will confound subsequent analyses depends on the batch--group balance. We have reanalysed two cases with varying degree of unbalance. The first case represents a ``worst case'' scenario since it both has a very unbalanced batch--group design, and was in part selected since it provided a methods description which allowed us to assess the use of ComBat, and perform a full re-analysis of the data. The second was taken from the original publication of the ComBat method (\citealp{Johnson2007}).

@@

In an experiment described in \citet{Towfic2014}, the effect of glatiramer acetate (a medicine for multiple sclerosis) was compared to the effect of a generic drug. Cells were treated with glatiramer acetate (34 samples), the generic drug (11 samples), or one of 14 other treatments (64 samples), and mRNA was measured using microarrays. A batch effect correlating to the chip (Illumina WG-6\_V2, six samples per chip, 17 chips in total) was observed and adjusted for with ComBat using treatment as a covariate. Batch adjusted data were then tested for differentially expressed genes, yielding hundreds of differentially expressed probes (Table S5, \citealp{Towfic2014}) using a $\text{FDR}<0.05$ threshold. Unfortunately, the batch--treatment design was highly unbalanced, with several batches having only one of the main treatments of interest. We re-analyzed the cited data (GEO: GSE40566) without using ComBat, but instead blocked for batch effect in \texttt{limma}, detecting only 9 differentially expressed genes at $\text{FDR}<0.05$. The sanity check  outlined above with random numbers was also carried out. The distribution of p-values for different settings are shown in Figure~\ref{fig:comparison}a.

\NB{Einars tekst:} It should be noted that the original analyses by \citet{Towfic2014} differed from our re-analysis in one critical manner which was clarified upon contacting the authors. Their analyses used a different preprocessing of the microarray data (GEO: GSE61901) in which two technical replicates of each biological sample was included, corresponding to the two different slots of the bead-microarray. They refer to \citet{Bakshi2013} for analyses where the two slots have been combined to provide one set of expression values per sample.

\NB{Vegards tekst:} Our conclusion is that most of the genes found as differentially expressed in the analysis described in \citet{Towfic2014} are false positives. This example represents somewhat of a ``worst case'' scenario for applying ComBat, since it both has a very unbalanced batch--group design and an a priori assumption of no difference. Our re-analysis, although based on the "methods" section of Towfic et al, deviates in important ways from the analysis that was actually performed in the paper (personal communication, Ben Zeskind). More detailed information on this aspect can be found in the extended github repostiory.

@@

A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were applied, i.e batch labels or whether group labels were supplied as covariates. Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches. The incorporation of ComBat into various analysis tools appears to make it more accessible, but its use less transparent. Such tools include GenePattern (\citealp{Reich2006}, version 3.9.0), AltAnalyze (\citealp{Emig2010}, version 2.0.8) and SCAN.UPC (\citealp{Piccolo2013}, version 2.6.3), which offers ComBat with covariates, as well as TCGA and  inSilicoMerging (\citealp{ Taminau2012}, version 1.8.7), presently without covariates as an option. Especially troublesome, are the ComBat implementations in ChAMP (\citealp{Morris2014}, version 1.2.8) and intCor (\citealp{Fernandez-Albert2014}, version 1.03), where the use of covariates seems automatic and hidden from the user in the method call, while there is no mention of covariates in the documentation or articles.

@@

\bibliographystyle{biorefs}
%\input{manuscript.bbl} % Include bibliography bbl-file
\bibliography{D:/Projects/_Appl_/_TeX_/library} % To use with BibTeX to generate bbl-file
%\bibliography{refs}

@@

\begin{figure}[!p]
\begin{center}$
\begin{array}{ccc}
\includegraphics[width=4.5cm]{Fig/leekrandomdatapvalues.pdf} &
\includegraphics[width=4.5cm]{Fig/leekqqplot.png}&
\includegraphics[width=4.5cm]{Fig/samplesizescaling_ComBat.pdf}
\end{array}$
\end{center}
\caption
{
This is a sanity check where the recommended use of ComBat fails, adapted from the user guide in the \texttt{sva} package. Real data are substituted with random numbers from a normal distribution, but the batch--group design is retained. ComBat is applied, followed by an F-test.
a) P-value distribution
b) QQ plot of the F-statistics
c) P-value distributions for 3 equally unbalanced random number experiments with different sample sizes, 12,120 and 1200 samples from two study groups with a 1:5 and 5:1 distribution in two batches. A random batch effect is added for 10\% of the 20000 genes. This example is not from the \texttt{sva} package.
}
\label{fig:sanity}
\end{figure}
